---
title: 'Day 2 Part 1: Dive!'
author: "Zack Wixom"
date: "12/3/2021"
output: github_document
---

```{r setup}
# Packages
library(tidyverse)
library(runner)
library(zoo)

# Read in data
data_raw <- read_csv(here::here("day2", "data", "day2_data_raw.csv"))
```

### The Prompt

Now, you need to figure out how to pilot this thing.

It seems like the submarine can take a series of commands like forward 1, down 2, or up 3:

- forward X increases the horizontal position by X units.
- down X increases the depth by X units.
- up X decreases the depth by X units.

*Note that since you're on a submarine, down and up affect your depth, and so they have the opposite result of what you might expect.*

The submarine seems to already have a planned course (your puzzle input). You should probably figure out where it's going. For example:

```
forward 5
down 5
forward 8
up 3
down 8
forward 2
```

**Your horizontal position and depth both start at 0**. The steps above would then modify them as follows:

- `forward 5` adds 5 to your horizontal position, a total of 5.
- `down 5` adds 5 to your depth, resulting in a value of 5.
- `forward 8` adds 8 to your horizontal position, a total of 13.
- `up 3` decreases your depth by 3, resulting in a value of 2.
- `down 8` adds 8 to your depth, resulting in a value of 10.
- `forward 2` adds 2 to your horizontal position, a total of 15.

After following these instructions, you would have a horizontal position of 15 and a depth of 10. (Multiplying these together produces 150.)

Calculate the horizontal position and depth you would have after following the planned course. 

### The Challenge

First I need to separate the data between the direction and the movement. as well as add depth and horizontal position.

```{r clean}
data <- data_raw %>% 
  separate(col = x, into = c("direction", "movement")) %>% 
  mutate(
    position = rep(0),
    depth = rep(0)
  )

data$movement <- as.numeric(data$movement)

data

```
Sweet, now we need to use this to calculate our position and depth. We start at (0,0) and we need to see where we are at by 2000 rows.

```{r}
sub_position <- data %>% 
  mutate(
    position = case_when(
      direction == "forward" ~ movement,
      direction == "down" ~ 0,
      direction == "up" ~ 0
    ),
    depth = case_when(
      direction == "forward" ~ 0,
      direction == "down" ~ movement,
      direction == "up" ~ -movement
    )
  )
  

```

Now I need to sum it all up and I should have my position:

```{r}
# Sumarize final depth and position
sub_position %>% 
  summarize(
    position_final = sum(position),
    depth_fianl = sum(depth)
  )

# multiply final position
sum(sub_position$position)*sum(sub_position$depth)
  

```


### The Answer

**What do you get if you multiply your final horizontal position by your final depth?**

1,746,616