---
title: 'Day 1 Part 2: Three-Measure Sliding Window'
author: "Zack Wixom"
date: "12/3/2021"
output: github_document
---

```{r setup}
# Packages
library(tidyverse)
library(runner)
library(zoo)

# Read in data
data <- read_csv(here::here("day1", "data", "day1_data.csv"))
```

### Prompt

Considering every single measurement isn't as useful as you expected: *there's just too much noise in the data.*

Instead, consider sums of a **three-measurement sliding window**. Again considering the above example:

```
199  A      
200  A B    
208  A B C  
210    B C D
200  E   C D
207  E F   D
240  E F G  
269    F G H
260      G H
263        H

```

Start by comparing the first and second three-measurement windows. The measurements in the first window are marked A (199, 200, 208); their sum is `199 + 200 + 208 = 607`. The second window is marked B (200, 208, 210); its sum is `618`. The sum of measurements in the second window is larger than the sum of the first, so this first comparison increased.

Your goal now is to count the number of times the sum of measurements in this sliding window increases from the previous sum. So, compare A with B, then compare B with C, then C with D, and so on. Stop when there aren't enough measurements left to create a new three-measurement sum.

In the above example, the sum of each three-measurement window is as follows:

```

A: 607 (N/A - no previous sum)
B: 618 (increased)
C: 618 (no change)
D: 617 (decreased)
E: 647 (increased)
F: 716 (increased)
G: 769 (increased)
H: 792 (increased)

```
In this example, there are 5 sums that are larger than the previous sum.


### The Challenge

First I need to figure out how to add these three-measure sliding windows to my data so I can calculate the answer.

I am going to use the `zoo` package which allows me to use `roll` functions. So I can apply functions to windows within the data

Also we are using the same data from day1 but I am going to first make sure I can replicate the example shown above before.

```{r prep}
# Test based on example given
B <- c(199, 200, 208, 210, 200, 207, 240, 269, 260, 263)

# calculate
k <- 3
rollapply(B, 3, function(x) max(rollsum(x, k)), partial = TRUE, fill = NA)

# empty vector to fill
x <- vector(mode = "numeric", length = 10L)
# For Loop
for (i in seq_along(B)) {
  # print progress
  print(i)
  # Calculate rolling sum
  x[i] <- rollsum(B[i:(i+2)], k = 3)
}
# Convert to dataset
output <- as.tibble(x)

```

Now that it works, I can apply it to the whole data input.

```{r for loop}

# Empty Vector to fill
x <- vector(mode = "numeric", length = nrow(data))
# For Loop
for (i in seq_along(x)) {
  # Extract data we need
  y <- data[2]
  # Calculate rolling sum
  x[i] <- rollsum(y[i:(i+2),], k = 3)
}
# Convert to dataset
output <- as.tibble(x)

output

```

Sweet now that we have the calculations I'll join it back to the original and using similar method as `day1_code.Rmd` I'll calculate all the increases.

```{r summarize}
# join columns
data1 <- cbind(data, output)

# calculate increase/decrease
windows <- data1 %>% 
  mutate(
    diff = value - lag(value, default = value[1])
  ) %>% 
  mutate(
    delta = case_when(
      diff == 0 ~ "No change",
      diff < 0 ~ "Decrease",
      diff > 0 ~ "Increase"
    )
  )

# Summarize
windows %>% 
  group_by(delta) %>% 
  summarize(
    n = n()
  )

# write salutation data to csv
write.csv(windows, here::here("day1", "data", "day1_windows.csv"))

```

### The Answer

**How many sums are larger than the previous sum?**

1103


