---
title: "Day 1 Part 1: Sonar Sweep"
author: "Zack Wixom"
date: "12/2/2021"
output: github_document
---

```{r setup}
# Packages
library(tidyverser)

```

### Prompt

As the submarine drops below the surface of the ocean, it automatically performs a sonar sweep of the nearby sea floor. On a small screen, the sonar sweep report (your puzzle input) appears: each line is a measurement of the sea floor depth as the sweep looks further and further away from the submarine.

For example, suppose you had the following report:

```
199
200 
208
210
200
207
250
```

This report indicates that, scanning outward from the submarine, the sonar sweep found depths of *199*, *200*, *208*, *210*, and so on.

The first order of business is to figure out how quickly the depth increases, just so you know what you're dealing with - you never know if the keys will get carried into deeper water by an ocean current or a fish or something.

To do this, count the number of times a depth measurement increases from the previous measurement. (There is no measurement before the first measurement.) In the example above, the changes are as follows:

```
199 (N/A - no previous measurement)
200 (increased)
208 (increased)
210 (increased)
200 (decreased)
207 (increased)
240 (increased)
269 (increased)
260 (decreased)
263 (increased)
```

In this example, there are 7 measurements that are larger than the previous measurement.

### The Challenge

**Import input data**

First I need the input data that I will get from the [source](https://adventofcode.com/2021/day/1/input) and convert to a data set in `day1_data.R`

```{r}
data <- read_csv(here::here("day1", "data", "day1_data.csv"))

head(data)

```

**Count Depth Measurement Increases**

```{r}
sonar_sweep <- data %>% 
  mutate(
    diff = x - lag(x, default = x[1])
  ) %>% 
  mutate(
    delta = case_when(
      diff == 0 ~ "No change",
      diff < 0 ~ "Decrease",
      diff > 0 ~ "Increase"
    )
  )

sonar_sweep %>% 
  group_by(delta) %>% 
  summarize(
    n = n()
  )

write.csv(sonar_sweep, here::here("day1", "data", "day1_sonar.csv"))
```


### The Answer

**How many measurements are larger than the previous measurement?**

1139


